---
title: "Line plot"
author: "Sue McClatchy"
date: "July 14, 2016"
output: html_document
---

## Learning Objectives
1. Load data from a URL.
2. Convert date objects and formats.
3. Log-transform the scale of a plot.
4. Specify breaks in the x and y axes.
5. Use plot annotations, custom axis labels, and axis themes.
6. Save plot as a PDF.

## Introduction
Time series plots are valuable but can be tricky to create because date and time formats in computing are not straightforward. Here we'll recreate the plot showing growth in the [Sequence Read Archive](http://www.ncbi.nlm.nih.gov/Traces/sra/) from 2008 to present.

Load the ggplot and scales libraries. You'll need to install the packages
first if you haven't done so already. Install them from the Packages tab,
or use the install.packages() command. Use double quotes around the package
name.
```{r eval = FALSE}
install.packages("ggplot2")
install.packages("scales")
```
You only need to install a package once to download it into your machine's library. Once you have installed the package on your machine, you need to load the library in order to use the functions contained in the package.
```{r}
library(ggplot2)
library(scales)
```
When you load a library you'll get a warning message indicating the R version in which the library was built. If it's different from the R version that you're running, you might occasionally run into problems depending on the library and the functions it contains. To find out what version of R you have, type

```{r}
version
```

The version of R is given as version.string, followed by the nickname for the version.

Read in the [Sequence Read Archive](http://www.ncbi.nlm.nih.gov/Traces/sra/sra_stat.cgi) database growth file  and view structure.

```{r}
sra <- read.csv("http://www.ncbi.nlm.nih.gov/Traces/sra/sra_stat.cgi")
str(sra)
```

The data have `r nrow(sra)` rows and `r ncol(sra)` columns. The first column is listed as a factor when in fact it is a date in the MM/DD/YYYY format. A factor is a categorical variable (i.e. red, green, blue, or low, middle, and high-income). Date variables are a data type that includes month, day and year, and that have their own specific function to extract weekdays or count the number of days until an event, for example. To place dates on the x-axis in proper order, convert the first column to a date object in the YYYY-MM-DD format. First check to make sure that the first several dates will be converted correctly.

```{r}
class(sra$date)
class(as.Date(sra$date, format = "%m/%d/%Y"))
head(sra$date)
head(as.Date(sra$date, format = "%m/%d/%Y"))
```

The first several dates convert correctly from MM/DD/YYYY to YYYY-MM-DD format, which is the default date format in R and the International Standards Organization (ISO) standard date format. Now convert all dates to the new format, and convert them from a factor variable to a date variable.

```{r}
sra$date <- as.Date(sra$date, format = "%m/%d/%Y")
class(sra$date)
head(sra$date)
str(sra)
```

Plot the number of bases and open access bases in SRA by layering lines over the time series on the x-axis. Use the ggplot() function, which is found in the ggplot2 library. Quick reminder of ggplot syntax: ggplot(data, mapping) + layer(). 
Start by plotting bases alone on the y-axis.
```{r}
ggplot(data = sra, mapping = aes(date)) + 
  geom_line(aes(y = bases))
```

Can you identify the ggplot syntax in the code above? Remember: ggplot(data, mapping) + layer(). The layer is a line.
Now add a second layer, a line representing growth of open access bases. Use the up arrow on your keyboard to retrieve the last bit of code you wrote, and add onto that.
```{r}
ggplot(data = sra, mapping = aes(date)) + 
  geom_line(aes(y = bases)) + 
  geom_line(aes(y = open_access_bases))
```

Change the line colors and sizes. Use the up arrow to retrieve the last code you ran, and add onto it rather than typing anew. More typing = more pain and suffering.

```{r}
ggplot(data = sra, mapping = aes(date)) + 
  geom_line(aes(y = bases), colour = "blue", size = 1.5) + 
  geom_line(aes(y = open_access_bases), colour = "yellow", size = 1.5)
```

The plot seems to show zero bases until the year 2010. Check the smallest number of total bases and open access bases in the data. Are the minimum numbers of bases both zero?

```{r}
min(sra$bases)
min(sra$open_access_bases)
max(sra$bases)
max(sra$open_access_bases)
```

The smallest number of bases is `r format(min(sra$bases), digits = 3)`, and the largest number is `r format(max(sra$bases), digits = 3)`. The smallest number of open access bases is `r format(min(sra$open_access_bases), digits = 3)`, and the largest number is `r format(max(sra$open_access_bases), digits = 3)`. We need to transform the y axis to logarithmic for accurate display, so that the plot doesn't show values of zero that don't exist.
```{r}
ggplot(data = sra, mapping = aes(date)) + 
  geom_line(aes(y = bases), colour = "blue", size = 1.5) + 
  geom_line(aes(y = open_access_bases), colour = "yellow", size = 1.5) +
  scale_y_log10()
```

Now manually define the y axis breaks so that each order of magnitude is represented.
```{r}
ggplot(data = sra, mapping = aes(date)) + 
  geom_line(aes(y = bases), colour = "blue", size = 1.5) + 
  geom_line(aes(y = open_access_bases), colour = "yellow", size = 1.5) +
  scale_y_log10(breaks=c(1e+10, 1e+11, 1e+12, 1e+13, 1e+14, 1e+15))
```

Supply simpler superscripted exponents on the y axis labels. These are easier to read and interpret than are combinations of digits, plus signs, and the letter e.
```{r}
ggplot(data = sra, mapping = aes(date)) + 
  geom_line(aes(y = bases), colour = "blue", size = 1.5) + 
  geom_line(aes(y = open_access_bases), colour = "yellow", size = 1.5) +
  scale_y_log10(breaks=c(1e+10, 1e+11, 1e+12, 1e+13, 1e+14, 1e+15),
                labels = expression("10"^"10", "10"^"11", "10"^"12",
                                    "10"^"13", "10"^"14", "10"^"15"))
```

Compare to the plot at the [Sequence Read Archive](http://www.ncbi.nlm.nih.gov/Traces/sra/).
Notice the difference in x and y axis labels and starting points. Our plot starts in the year 2000 with `r format(min(sra$open_access_bases), digits = 3)` bases. The [Sequence Read Archive](http://www.ncbi.nlm.nih.gov/Traces/sra/) plot starts before the year 2009 with 10^12^ bases.

How many data points are there in each year?

```{r}
table(format(sra$date, "%Y"))
```

Remove the single measurements in years 2000 and 2007. These two data points extend 
the line backward in time and give misleading information about the actual growth in the number of bases by making it appear as if data collection was sequential year-over-year. In fact, there were many years missing data altogether. Start at year 2008, the first year of sustained,  consecutive data collection.

```{r}
which(format(sra$date, "%Y")==c("2000", "2007"))
sra <- sra[-(1:2),]
table(format(sra$date, "%Y"))
```

Save the plot to a variable now that we've removed the two data points. To view the plot, type the name of the variable.
```{r}
logplot <- ggplot(data = sra, mapping = aes(date)) +
  geom_line(aes(y = bases), colour = "blue", size = 1.5) + 
  geom_line(aes(y = open_access_bases), colour = "yellow", size = 1.5) +
  scale_y_log10(breaks=c(1e+10, 1e+11, 1e+12, 1e+13, 1e+14, 1e+15),
                labels = expression("10"^"10", "10"^"11", "10"^"12", "10"^"13", "10"^"14",
                                    "10"^"15"))
logplot
```

Specify the breaks in the x-axis so that each year is shown. Redefine the variable
logplot with these changes.
```{r}
logplot <- logplot + scale_x_date(labels = date_format("%Y"), breaks = date_breaks("years"))
logplot
```

Add text annotations to the plot to label the lines for total and open access bases.
```{r}
logplot <- logplot + 
  annotate("text", x=sra$date[1100], y=1.6e+15, label="total bases") + 
  annotate("text", x=sra$date[1700], y=1.25e+14, label="open access bases")
logplot

```

Add axis labels.
```{r}
logplot <- logplot + 
  xlab("Year") + 
  ylab("Number of bases")
logplot
```

Add a title. Redefine logplot each time to save the changes.
```{r}
logplot <- logplot + 
  ggtitle("Sequence Read Archive Database Growth")
logplot
```

Save the plot as a PDF using the pdf() command. You can also save as a png(), jpeg(), tiff(), or bmp() using the corresponding command. Provide a file name inside the parentheses surrounded by double quotes. Be sure to turn the graphics device off afterward to return graphics output to your plot window in RStudio. Use dev.off() to turn off the graphics device.
```{r}
pdf(file = "SRA-database-growth.pdf")
print(logplot)
dev.off()
```

Now locate the file on your machine and open it. Compare with the plot at the [Sequence Read Archive](http://www.ncbi.nlm.nih.gov/Traces/sra/).

> Code Challenge: Plot bytes and open access bytes. Change to log scale. Color the lines, add text annotations, label the axes and add a title. Save the plot to a variable and open the file to view your plot. 

